# Analyzing Data Variation {.unnumbered}

> *Variation is the tendency of the values of a variable to change between each measurement*

This variation can happen due to different reasons:
  
  * **Error** when measuring a constant quantity (ligth speed).
  * Measuring the same variable in **different subjects**.
  * Measuring the same variable for the same subject but **different moment**.
  * ...

We can see patterns of the variation, this information can be useful. We can understand these patterns visualizing the **distribution of the different values taken by the variable**. 


## Variation in categorical variables

Remember that we can visualize the distribution of a categorical variable as follows...


```{r, echo=TRUE, message=FALSE}
library(ggplot2)

# Number of gears for mtcars data set
ggplot(mtcars, aes(x=gear)) +
  geom_bar() +
  labs(x = "Number of gear")
```

Each bar of the last plot shows us the number of observations we can found for every value of the variable. We can obtain the number of observations with this simple code...

```{r, echo=TRUE, message=FALSE}
library(tidyverse)

mtcars %>% 
  group_by(gear) %>% 
  summarise(observations = n())
```


## Variation in continuous variables

For visualizing the distribution of continuous variables we can use the histogram...

```{r, echo=TRUE}
ggplot(data = iris, aes(x = Petal.Length)) +
  geom_histogram(bins = 20) +
  labs(x = "Petal Length (cm)")
```

## Analyzing Variation

When analyzing the plots that shows the data variation we need to focus in two things...

* Normal values
* Anomalies (outliers)

Questions like...

* Most common values?
* Lest common values?
* Strange patterns in the data?

We can see it with an example: *Petal Length Analysis*

```{r, echo=TRUE}
ggplot(data = iris, aes(x = Petal.Length)) +
  geom_histogram(bins = 20) +
  labs(x = "Petal Length (cm)")
```

The *clusters* of similar values makes us think there are subgroups in the data. For understanding it, we can ask...

* What have the observations of each group in common?
* What differences are between the observations of each group?
* How can we explain each subgroup?
* There are two groups... one with shorter petal length (<2cm) and other large (>2cm)
* Species variable can explain this separation?


We can analyze multiple histograms in the same plot, using colors for example.

```{r, echo=TRUE}
ggplot(data = iris, 
       aes(x = Petal.Length, fill = Species)) +
  geom_histogram(bins = 20) +
  labs(x = "Petal Length (cm)")
```

Other option, we can use ggplot geometric: `geom_freqpoly()` that works similar to the histogram but showing them with lines.

```{r, echo=TRUE}
ggplot(data = iris, 
       aes(x = Petal.Length, color = Species)) +
  geom_freqpoly() +
  labs(x = "Petal Length (cm)")
```


## Variation: Outliers

Find and analyze **outliers** have the same importance as analyze the normal values we can found. 

Why we can have outliers...

  * Sometimes... measuring error 
  * Others, further analyze is needed. 
  
We can see it with an example: *Diamonds Width Analysis*

We are going to analyze the distribution of the width of the diamonds in the data set **diamonds** inside R. Apparently all looks good right? (*Hint: Look at x axis?*)

```{r, echo=TRUE}
ggplot(diamonds, aes(x = y)) + 
  geom_histogram(binwidth = 0.5)
```

Looks that there is more space than the one we need to visualizing the data, or probably there is some data we are not able to see, a bit of *zoom* may help...

```{r, echo=TRUE}
ggplot(diamonds, aes(x=y)) +
  geom_histogram(binwidth = 0.5) +
  coord_cartesian( ylim = c(0,50) )
```

Looks there are some outliers near 0, 30 and 60. We can now go directly to the data and **filter** to extract them..

```{r, echo=TRUE}
diamonds %>% 
  filter( y < 3 | y > 20)
```

**Conclusions found after analysis**

* Looks there are diamonds with width 0, maybe incorrect data...

* We have detected diamonds with a very big size, but not too expensive...

## Actions with outliers

* Try to do the analysis with and without outlers, this will help us identify the impact they have. 

Depending on the impact effect...

  * Minimum effect and we don't know the origin... replace them as `NA` values. 
  
  * Otherwise... eliminate them **justifying** the reason and document / inform their elimination from the analysis. 
  

### Droping outliers

When we need to drop some outliers from the data, we have two options...

* Eliminate the whole row (big loss of information)

```r
diamonds2 <- diamonds %>%
  filter(y>=3 & y<=20)
```

* Replace them by `NA` (missing value)

```{r, echo=TRUE}
diamonds2 <- diamonds %>%
  mutate(y = if_else(y<3 | y>20, NA, y))
```

::: callout-tip

# if_else

Conditional function to use inside mutate, first argument the condition, then if TRUE otherwise FALSE in the last argument. 

:::


## Missing Values

When missing values are present in the data we will see a warning in the plot...

```{r, echo=TRUE, warning=TRUE}
ggplot(data=diamonds2, 
       mapping = aes(x=x, y=y)) + 
  geom_point()
```

We can quickly eliminate them from the visualization using `na.rm = T` argument...

```{r, echo=TRUE}
ggplot(data=diamonds2, 
       mapping = aes(x=x, y=y)) + 
  geom_point(na.rm = T)
```


