# Analyzing Covariance {.unnumbered}

```{r, echo=F,message=F}
library(tidyverse)
```


> *Covariance is the tendency between two or more values to change simultaneously.*

As we have seen, for analyzing covariance we will need to understand the type of variables...

* Categorical - Continuous
* Categorical - Categorical
* Continuous - Continuous

## Covariance: Categorical - Continuous

We can use the `geom_freqpoly` geometry to see how the continuous variable relates with the each category. 

```{r, echo=TRUE, message=FALSE}
ggplot(data = diamonds, aes(x = price, color = cut)) +
  geom_freqpoly()
```

Other option with density plots. `..density..` will paint the density in the y axis, that is the normalized count to make the area of each polygon 1.

```{r, echo=TRUE, message=FALSE}
ggplot(data = diamonds, 
       aes(x = price, y = ..density.., color = cut)) +
  geom_freqpoly()
```

Using boxplots to analyze too...

```{r, echo=TRUE}
ggplot(data = mpg, aes(x = class, y = hwy)) +
  geom_boxplot()
```

Reorder it os is more clean...

```{r, echo=TRUE}
#| code-line-numbers: "2,3" 
ggplot(data = mpg, 
       aes(x = reorder(class, hwy, FUN = median),
           y = hwy)) +
  geom_boxplot()
```


## Covariance: Categorical - Categorical
We have seen bar plots, but more ways are possible...

```{r, echo=TRUE}
ggplot(diamonds, aes(x=cut, y=color)) +
  geom_count()
```


Let's do it with code using **dplyr** (two ways)

```{r, echo=TRUE}
diamonds %>% 
  group_by(color, cut) %>% 
  summarise(n = n())
```

More visual approach, using color mapping...

```{r, echo=TRUE}
diamonds %>% 
  count(color, cut) %>% 
  ggplot(aes(x = color, y = cut, fill = n)) +
  geom_tile()
```

## Covariance: Continuous - Continuous

Normally we will use an scatter plot with `geom_point()`, but we can find some problems when there are too many data points in big datasets, they will overalp each other. 

In these cases we will need to use `alpha` aesthetic:

```{r, echo=TRUE}
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point(alpha = 0.01)
```

Another approach...

```{r, echo=TRUE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_bin2d()
```

Another approach... but better 

```{r, echo=TRUE}
# install.packages("hexbin")
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_hex()
```

Box plots work here too:

```{r, echo=TRUE, message=FALSE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_boxplot(aes(group = cut_width(carat, 0.1)))
```

Same number of point for each bin:


```{r, echo=TRUE, message=FALSE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_boxplot(aes(group = cut_number(carat, 20)))
```


## Patterns and models

**Patterns** reveal hints regarding relations between variables. When we see one we must analyze...

* Can be random?
* How we can describe the relation that shows?
* Is relevant?
* Are there any other variables affected by this relation?
* This relation changes if we analyze subgroups of data?

**Models** can be used for extract patterns of the data, like eliminating the effect of a variable and study the remaining effect. 