# Parsing Data {.unnumbered}

```{r, echo=FALSE, message=FALSE}
#| code-line-numbers: false
library(tidyverse)
```


**read_csv** function realizes a parsing of text file. This means it converts a text file in an organized data structure.

## Parsing vectors

For understanding better who it is does that we need to understand **parsing** functions. This functions will transform a vector of characters into an speciallized data vector.

```{r, echo=TRUE}
#| code-line-numbers: false
str(parse_logical(c("T", "F", "F")))

str(parse_integer(c("1", "4")))

str(parse_date(c("2022-01-04")))
```

All `parse` functions are uniform: the first argument is the vector to parse and the `na` argument lend us decide with value treat as a missing value. 

```{r, echo=TRUE}
#| code-line-numbers: false
parse_integer(c("1", "4", "."), na=".")
```

If something goes wrong during the process a warning message will be displayed to inform us

```{r, echo=TRUE, warning=TRUE}
#| code-line-numbers: false
parse_integer(c("1", "4", "c"))
```

## Kind of parsers

* `parse_logical()` and `parse_integer()` for logical and integer values.

* `parse_double()` and `parse_number()` for numbers (second one more flexible)

* `parse_factor()` for categorical values (factors)

* `parse_datetime()`, `parse_date()` and `parse_time()` for times and dates.

## Parsing numbers

Numbers are quite complex as they have three main problems:

1) Numbers can be described differently depending on the culture/language: 1,46 or 1.45

```{r, echo=TRUE}
#| code-line-numbers: false
parse_double("1,45", locale = locale(decimal_mark = ","))
```

2) Sometimes we can find symbols in the numbers: 100%

```{r, echo=TRUE}
#| code-line-numbers: false
parse_number("100$")

parse_number("39%")

parse_number("Me costó 100 €")
```

3) Related with the first issue, representation can change on aggregation level: 1.000.000 or 1000000 or 1,000,000 or 1000K or 1M 

```{r, echo=TRUE}
#| code-line-numbers: false
parse_number("%100,330")

parse_number("100.330€$", locale = locale(grouping_mark = "."))
```

## Parsing Strings

Computers represent text using hexadecimal code (look for [ASCII](https://elcodigoascii.com.ar))

```{r, echo=T}
#| code-line-numbers: false
charToRaw("Hello")
```

The mapping between hexadecimal values and characters are called **encoding**. By default **readr** uses UTF-8, general by default. 

Sometimes in cal fail with old systems files:

```{r, echo=T}
#| code-line-numbers: false
x1 <- "Nos vemos ma\xf1ana"
parse_character(x1, locale = locale(encoding = "ISO-8859-1"))
```

We can avoid issues, analyzing the text prior to find the best encoding

```{r, echo=T}
#| code-line-numbers: false
x1 <- "Nos vemos ma\xf1ana"
guess_encoding(charToRaw(x1))
```

## Parsing Factors

n this case we will need to provide ```parse_factor()``` a vector with the possible levels:

```{r, echo=T}
#| code-line-numbers: false
parse_factor(
  c("man", "woman", "woman", "unicorn"),
  levels = c("man", "woman")
)
```

## Parsing Dates and Times

`parse_datetime()` for dates with hour. In the following order: *year, month, day, hour, minute and second*

```{r, echo=T}
#| code-line-numbers: false
parse_datetime("2024-01-01T1015")
```

`parse_date()` for dates. In the following order: *year, month, day* using `-` or `/` as separator

```{r, echo=T}
#| code-line-numbers: false
parse_date("2024/01/01")
```

`parse_time()` for time. In the following order: *hour, minute, seconds* using `:` as separator. `am` or `pm` are optional.

```{r, echo=T}
#| code-line-numbers: false
parse_time("10:15 am")
```

You can define a format of the input to help the parsing...
```{r, echo=T}
#| code-line-numbers: false
parse_date("01/04/2023", "%d/%m/%y")
```

::: {layout-ncol="3" style="font-size:20px;"}

#### Year

* **%Y** 4 digits

* **%y** 2 digits*

#### Month

* **%m** 2 digits

* **%b** "jan"

* **%B** "january"

#### Day

* **%d** 2 digits

* **%e** space

::::

*00-69 -> 2000-2069, 70-99 -> 1970-1999

You can define a format of the input to help the parsing...
```{r, echo=T}
#| code-line-numbers: false
parse_time("10-30", "%H-%M")
```

::: {layout-ncol="2" style="font-size:20px;"}


#### Time

* **%H** 0-23 hours (24)

* **%p** 0-12 hours (12), use with **%p**

* **%p** AM/PM

* **%M** minutes


#### __

* **%s** seconds (integer)

* **%OS** real seconds

* **%Z** time zone (e.g. America/Chicago)

* **%z** UTC offset (e.g. +0800)

::::

Discover more about formatting dates [here](https://readr.tidyverse.org/reference/parse_datetime.html)

## Parsing a file

### How readr parse correctly the data?

**readr** will guess the correct the type of each column as follows:

* Read the first 1000 rows

* Guess the type using a determined heuristic

```{r, echo=T}
#| code-line-numbers: false
guess_parser("2022-01-01")

str(parse_guess("2022-01-01"))
```

Sometimes it can fail...

* It is possible that the first 1000 rows have special cases (e.g. 1000 integer values in a double variable)

* Missing values can appear at the column. 

It is suggested to try to inform the type of each column when it is possible. 

```{r, echo=TRUE, eval=FALSE}
#| code-line-numbers: false
# Not specifying the types
read_csv("data/df-na.csv")
```

We can force it two ways...

```{r, echo=TRUE, eval=FALSE}
#| code-line-numbers: false
# Specifying the types
read_csv("data/df-na.csv",
         col_types = list(col_double(),
                          col_character(),
                          col_character()))
```

```{r, echo=TRUE, eval=FALSE}
#| code-line-numbers: false
# Other alternative
read_csv("data/df-na.csv", col_types="ncc")
```