# Pivots & Columns {.unnumbered}

```{r, echo=FALSE, message=FALSE}
#| code-line-numbers: false
library(tidyverse)
```

## Pivoting Data
Normally, the data we receive is not properly organized. So the first, step is to identify the variables and observations. 

Once identified, we will need to solve one of these problems:

* A variable is disperse in several columns.
* An observation is disperse in multiple rows. 

This can be resolved using `pivot_longer()` or `pivot_wider()` 

### `pivot_longer()`

The same variable is organized in different columns:

```{r}
#| code-line-numbers: false
table4a
```


We will need to **pivot** this columns a new pair of variables. We need to know:

* The columns to pivot. 
* The name of the variable that will receive the column names. `year` in this case.
* The name of the variable that will receive the values. `cases` in this case.


```{r, echo=TRUE}
table4a %>% 
  pivot_longer(
    cols = c(`1999`, `2000`), # columns to pivot
    names_to = "year", # variable to host colnames
    values_to = "cases" # variable to host values
  )

```

::: {.callout-note appearance="simple" style="font-size:30px;"}

Take into account we use (`) symbol in the name of the columns because they have a number as name. 

:::

### `pivot_wider()`

The same observation is organized in several rows. In this example each observation shows each country each year. You see each observation takes two rows...

```{r}
#| code-line-numbers: false
table2
```

We will need to organize the data showing only one row per observation. For doing that we need...

* The column where we will extract the names for the new variables. `type` in this case.

* The column where we will extract the valus for each variable. `count` in this case.

```{r, echo=TRUE}
table2 %>% 
  pivot_wider(
    names_from = type, # origin column with variables
    values_from = count # origin column with values
  )
```


### To Sum Up

* `pivot_longer()` will eliminate columns and create rows, makes data more large. 

* `pivot_wider()` will eliminate rows and create columns, makes data more wide. 

## Arranging Columns

### `separate`

We can found situations when a column contains information related to two or more variables.

```{r}
#| code-line-numbers: false
table3
```

We can use `separate()` to divide one column in two. 

```{r, echo=T}
table3 %>% 
  separate(rate, 
           into = c("cases", "population"))
```

By default it makes the separation when finds a non-alphanumeric value. But you can specify the character to use.

```{r, echo=T}
table3 %>% 
  separate(rate, 
           into = c("cases", "population"),
           sep = "/")
```

You can do it with integer values too as position to make separation.

```{r, echo=T, eval=F}
table5x <- table3 %>% 
  separate(year, 
           into = c("century", "year"),
           sep = 2 # make separation after 2 characters
           )
```

```{r}
table5x <- table3 %>% 
  separate(year, 
           into = c("century", "year"),
           sep = 2 # make separation after 2 characters
           ) %>% 
  select(-rate)
```


It is convenient to convert the value, otherwise the new columns will be Strings. 

```{r, echo=T}
table3 %>% 
  separate(rate, 
           into = c("cases", "population"),
           convert = TRUE)
```

### `unite`

For concatenate columns we can use `unite()`. It has as arguments the new variable, columns and separator to use.

```{r,echo=TRUE}
# Separated century and year
table5x %>% 
  select(country, 
         century, year)
```

```{r,echo=TRUE}
table5x %>% 
  unite(full_year, 
        century, year, 
        sep = "")
```

