# Dates & Time {.unnumbered}

As shown, for working with dates and times we are going to use `lubridate` package. It is now a part of tidyverse and is included when you load it, but in older R versions could not been included and you will need to load it individually.

```{r, echo=TRUE}
#| code-line-numbers: false
#| message: false
library(tidyverse) # use library(lubridate) for this package only
```

We can see three data types...

-   Dates
-   Time
-   Datetime, combination of the first two identifying a time period. Take into account that in tibbles they are identified as datetime (dttm) but R base uses POSIXct as normal type.

We can get today`s date...

```{r}
#| code-line-numbers: false
today()
```

Or current date time...

```{r}
#| code-line-numbers: false
now()
```

## Parsing dates

Normally we will receive this data as Strings, this will make us think use the parsing functions we learn at the first block of this unit, but...

`lubridate` has a more easy way of doing it with multiple purpose functions where we just need to specify the order with **y**, **m** or **d**

```{r, echo=TRUE}
#| code-line-numbers: false
ymd("2022-05-04")

ymd("22-May-04")

```

More examples, the syntax offered by `lubridate` has a lot of possibilites to parse dates and times:

```{r, echo=TRUE}
#| code-line-numbers: false
mdy("May 4th, 2022")

dmy("4 of May of 2022")

ymd_hms("2022-05-04 20:11:58")

mdy_hm("04/05/2022 00:01")
```

## `make_datetime()`

Some times we can receive the data in separate variables...

```{r, echo=TRUE}
library(nycflights13)
flights %>% 
  select(year, month, day, hour, minute)
```

We can simply create date with individual variables using `make_datetime()`.

```{r, echo=TRUE}
flights %>% 
  select(year, month, day, hour, minute) %>% 
  mutate(departure = make_datetime(year, month, day, hour, minute)) %>% 
  head(5)
```

## Change formats

You can change between formats...

```{r, echo=TRUE}
#| code-line-numbers: false
as_datetime(today())

as_date(now())
```

## Extracting components

You can extract components for dates and time...

```{r, echo=TRUE}
#| code-line-numbers: false
datetime <- ymd_hms("2022-05-04 12:34:56")
year(datetime)

month(datetime)

mday(datetime) # day of month

yday(datetime) # day of year

wday(datetime) # day of week

hour(datetime)

minute(datetime)

second(datetime)

```

In these examples are extractions are numerical. So we have the day, hour etc. But for some date attributes (like day of week or month) we have a name.  We change the behavior of the functions to obtain the label. 

```{r, echo=TRUE}
#| code-line-numbers: false
datetime <- ymd_hms("2022-05-04 12:34:56")
month(datetime, label = TRUE)

wday(datetime, label = TRUE, abbr = FALSE)

```

You can change labels output language.
```{r, echo=TRUE}
#| code-line-numbers: false
wday(datetime, label = TRUE, abbr = FALSE, locale = Sys.setlocale("LC_TIME", "es_ES"))
```

## Modifying components

We can modify the components using the same functions

```{r, echo=TRUE}
#| code-line-numbers: false
datetime <- ymd_hms("2022-05-04 12:34:56")

year(datetime) <- 2023
datetime

month(datetime) <- 01
datetime

hour(datetime) <- hour(datetime) + 1
datetime
```

## Computations with dates and time

For computing with dates we need...

-   **Duration**, exact number of seconds
-   **Periods**, other units like months and weeks.

### Duration

When making the difference between two dates, will obtain a `difftime` object...

```{r, echo=TRUE}
#| code-line-numbers: false
my_age <- today() - ymd("1998 Mar 4")
my_age

class(my_age)
```

This object will contain the information regarding time period in seconds, minutes, hours, days, etc... pretty ambiguous.

Alternative, we can work with **duration** (in seconds)

```{r, echo=TRUE}
#| code-line-numbers: false
as.duration(my_age)
```

Other useful functions... in seconds

```{r, echo=TRUE}
#| code-line-numbers: false
dseconds(15)

dminutes(10)

dhours(c(12, 24))

ddays(0:5)

dweeks(3)

dyears(1)
```

Making computations in seconds...

```{r, echo=TRUE}
#| code-line-numbers: false
dyears(1) + dweeks(12) + dhours(15)

2 + dweeks(12)

tomorrow <- today() + ddays(1)
last_year <- today() - dyears(1)
```

### Periods

Last methods will help making computations using seconds as our base... but we have other functions more intuitive for just operating with the exact period...

```{r, echo=TRUE}
#| code-line-numbers: false
seconds(15)

minutes(10)

hours(c(12, 24))

days(7)

```

```{r, echo=TRUE}
#| code-line-numbers: false
months(1:6)

weeks(3)

years(1)
```

Computing...

```{r, echo=TRUE}
#| code-line-numbers: false
ymd("2022-01-01") + years(1)

ymd("2022-01-01") + months(1)

ymd("2022-01-01") + days(1)
```

Take care when mixing both systems... or periods or duration not both...

```{r, echo=TRUE}
#| code-line-numbers: false
ymd("2022-01-01") + dyears(1)

ymd("2022-01-01") + years(1)

```
