# String {.unnumbered}

Remember an String is a text format here some examples...

```{r}
#| code-line-numbers: false
#| eval: false
"This is a String"
```

```{r}
#| code-line-numbers: false
#| eval: false
c("one", "two and three")
```

For doing all that we will use `stringr` a tidyverse package developed to work with Strings. All the functions in this package start with `str_`

```{r, echo=TRUE}
#| code-line-numbers: false
#| message: false
library(tidyverse) # use library(stringr) for this package only
```

## Basic string functions

### Length of a string

We can compute the length of an string using `str_length()`

```{r, echo=TRUE}
#| code-line-numbers: false
str_length("Hello")
```

All operations can be applied to vectors too...

```{r, echo=TRUE}
#| code-line-numbers: false
str_length(c("Hello", "World!"))
```

What will be the behavior with blank spaces? Blank spaces are consider characters too, so they will count on the total.

```{r}
#| code-line-numbers: false
str_length("CUNEF EDU")
```

### Subsetting strings

You can extract parts of strings...

```{r}
#| code-line-numbers: false
str_sub("Hello World",start = 1, end = 5)
```

Or by the end...

```{r}
#| code-line-numbers: false
str_sub("Hello World",start = -5, end = -1)
```

### Changing case

You can change the case of the letters. There are several options...

- All to lower case with `str_to_lower`
- All to upper case with `str_to_upper`
- All to title case with `str_to_title`, first letter of each word in upper case, rest on lower case.

```{r}
#| code-line-numbers: false
str_to_lower("HeLlo wOrlD")
```

```{r}
#| code-line-numbers: false
str_to_upper("HeLlo wOrlD")
```

```{r}
#| code-line-numbers: false
str_to_title("HeLlo wOrlD")
```

### Concatenating Strings

We can combine (concatenate) Strings using `str_c()`

```{r, echo=TRUE}
str_c("Va", "len", "cia")
```

You can specify the separation string used when doing the concatenation with the `sep` argument.

```{r, echo=TRUE}
str_c("One", "two", sep=",")
```

This works with vectors too...

```{r, echo=TRUE}
#| code-line-numbers: false
str_c("prefix-", c("X", "Y", "Z"), "-sufix")
```

For collapsing a full large text in a single string with `collapse` argument.

```{r, echo=TRUE}
str_c(c("A", "long", "time", "ago,", 
        "in", "a", "galaxy", "far", "far", "away..."), 
      collapse = " ")
```

## Regular Expressions

Regular expressions (*regexp*) are tools designed to find patterns in texts. We can do several thins with patterns...

- Find which strings complain with the pattern.
- Find in which position appears the pattern.
- Extract the pattern from the string.
- Replace the pattern in the string.

For example, we want to detect ES followed by two numbers"

```{r}
#| code-line-numbers: false
#| warning: false
x1 <- "ES49 "
x2 <- "Mi account number is ES34 30..."
x3 <- "I live in Spain (ES)"

str_view_all(c(x1,x2,x3), 'ES[0-9]{2}')
```

In the example we have created a pattern `'ES[0-9]{2}'` to match the structure we need to find 'ES' followed by two numbers. We do a search on the examples and we can see that the pattern has been found in the two first examples.

::: {.callout-note appearance="simple"}
In the example seen before, we have use `str_view`. This function is used to validate patterns and for learning purposes only.
:::

For building patterns, R has a wide range of tools we can use to develop the proper pattern syntax. General rules for building them are...

- exact word or text we want to find.
- `.` identifies any character
- `\\s` blank space
- `[]` set of elements e.g. `[abc]`: a or b or c
- `[0-9]` all numbers from 0 to 9
- `[a-z]` all letters in lower case
- `[A-Z]` all letters in upper case
- We can use `+` to repeat a pattern several times, e.g. two numbers: `[0-9][0-9]` or `[0-9]+` will work.
- `ab|bc` two possible valid options ab or bc
- `^` start pattern at the beginning
- `$` start pattern at the end

You test and try *regex* in the following [link](https://regexr.com) 

We will review all of them with function examples...

### Finding patterns: `str_detect`

For trying to find if a text complies with a pattern we can use `str_detect` function. 

Imagine that we want to find the words that contain `Le` (**exact match**) in the following list.

```{r}
x <- c("Cats", "Leia Organa", "Capital", "Leman")
str_detect(x, "Le")
```

The output of `str_detect` is a logical value. `TRUE` if the pattern appears and `FALSE` otherwise. 

So, how to use it properly on data analysis. Imagine the following example...

```{r}
sample_tb <- tibble(
  "texts" = c("234H",
                "B12, C15, E14",
                "Academic Year 1",
                "Two and a half Men",
                "Account ES14 2345 5678 2456")
)

sample_tb
```

We want to find all texts that contain a letter `a`. We will do it using `filter`...

```{r}
sample_tb %>% 
  filter(str_detect(texts, "a"))
```


The detect is only looking for exact match of letter `a`, let's try to find any one that starts with letter `a`

For that we need to use \$ symbol

```{r}
sample_tb %>% 
  filter(str_detect(texts, "^a"))
```

Looks nothing is returned, maybe is due to the case, we can solve it two ways...

```{r}
sample_tb %>% 
  filter(str_detect(texts, "^a|^A"))

sample_tb %>% 
  filter(
    str_detect(
    str_to_lower(texts), "^a"))
```
Last example, we want any text with a combination of two or more numbers...

```{r}
sample_tb %>% 
  filter(str_detect(texts, "[0-9][0-9]+"))
```

### Counting number of appearances: `str_count`

We can count the times the pattern is detected with `str_count`

For example, count the number of `a` in each text.

```{r}
sample_tb %>% 
  mutate(
    as = str_count(texts,"a")
  )
```

### Replacing: `str_replace`

Another usage of patterns is to detect them and make replacements, for example for treating data and replace it. We can do this with `str_replace()` or `str_replace_all()`.

In the example we want to replace the blank spaces by `_`\`

```{r}
sample_tb %>% 
  mutate(
    text_clean = str_replace(texts, "\\s", "_")
  )
```

By default it only replaces one match, if we want to apply it to every coincidence we need to use `str_replace_all`\`

```{r}
sample_tb %>% 
  mutate(
    text_clean = str_replace_all(texts, "\\s", "_")
  )
```

### Extracting: `str_extract`

We can also extract patterns matched...

In the example we extract any combination of a letter+number:

```{r}
sample_tb %>% 
  mutate(
   letter_num = str_extract(texts, "[A-Z][0-9]")
  )
```

